---
title: "Categorical Predictors and Dummy Variables"
author: "Osman Omer Mustafa"
date: "today"
format:
  html:
    toc: true
    code-fold: true
    theme: cosmo
    toc-location: left
---

## Load required libraries
```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(broom)
```

## Read data
```{r message=FALSE}
data <- read_csv("insurance.csv")
```

## Basic checks
```{r}
glimpse(data)
summary(data)
sum(is.na(data))
```

## Ensure categorical variables are factors
```{r}
data <- data %>% 
  mutate(
    sex = as.factor(sex),
    smoker = as.factor(smoker),
    region = as.factor(region)
  )

glimpse(data)
```

Categorical variables (sex, smoker, region) have been successfully converted to factors, 
enabling proper treatment in regression models. R will automatically create dummy variables for these predictors.

## Exploratory Data Analysis

### Summary statistics by sex
```{r}
summary_sex <- data %>% 
  group_by(sex) %>% 
  summarise(
    n = n(),
    mean_charges = mean(charges),
    sd_charges = sd(charges),
    median_charges = median(charges)
  )
summary_sex %>% knitr::kable(caption = "Insurance Charges by Sex")
```

Males and females show relatively similar average charges, with males slightly higher.
The difference appears modest compared to other factors like smoking status.

### Summary statistics by region
```{r}
summary_region <- data %>% 
  group_by(region) %>% 
  summarise(
    n = n(),
    mean_charges = mean(charges),
    sd_charges = sd(charges),
    median_charges = median(charges)
  )
summary_region %>% knitr::kable(caption = "Insurance Charges by Region")
```

Regional variation in charges is present but relatively small. 
The southeast region shows the highest average charges, while the southwest shows the lowest.

### Charges by sex
```{r}
ggplot(data, aes(x = sex, y = charges, fill = sex)) +
  geom_boxplot() +
  labs(
    x = "Sex",
    y = "Insurance Charges (USD)",
    title = "Insurance Charges by Sex"
  ) +
  theme(legend.position = "none") +
  theme(plot.title = element_text(hjust = 0.5))
```

The boxplots reveal overlapping distributions between males and females, 
with similar medians and ranges. Both groups contain outliers with exceptionally high charges.

### Charges by region
```{r}
ggplot(data, aes(x = region, y = charges, fill = region)) +
  geom_boxplot() +
  labs(
    x = "Region",
    y = "Insurance Charges (USD)",
    title = "Insurance Charges by Region"
  ) +
  theme(legend.position = "none") +
  theme(plot.title = element_text(hjust = 0.5))
```

Regional distributions are remarkably similar, with comparable medians and spread. 
This visual similarity suggests region may not be a strong predictor of insurance charges.

## Understanding Dummy Variables

### Check default contrasts
```{r}
contrasts(data$sex)
contrasts(data$region)
```

R uses treatment (dummy) coding by default. For sex, female is the reference category (coded as 0), 
while male is coded as 1. For region, northeast serves as the baseline, 
with the other three regions represented by dummy variables.

### Check levels order
```{r}
levels(data$sex)
levels(data$region)
```

The alphabetically first category becomes the baseline by default. 
This can be changed using the `relevel()` function if a different reference category is desired.

## Model 1: Sex Only

Fit a regression model using sex as the sole predictor to examine its relationship with insurance charges.

```{r}
model_sex <- lm(charges ~ sex, data = data)
```

### Model results
```{r}
tidy_sex <- tidy(model_sex, conf.int = TRUE)
tidy_sex %>% knitr::kable(caption = "Sex Model Coefficients")
```

The intercept represents average charges for females (baseline). 
The sexmale coefficient indicates males pay approximately $1,387 more than females on average. 
This difference is statistically significant (p < 0.001).

### Model performance
```{r}
glance_sex <- glance(model_sex)
glance_sex %>% knitr::kable(caption = "Sex Model Performance")
```

The model explains only about 0.5% of variance (R² ≈ 0.005), 
indicating sex alone is a weak predictor of insurance charges.

## Model 2: Region Only

Examine regional differences in insurance charges.

```{r}
model_region <- lm(charges ~ region, data = data)
```

### Model results
```{r}
tidy_region <- tidy(model_region, conf.int = TRUE)
tidy_region %>% knitr::kable(caption = "Region Model Coefficients")
```

The intercept represents average charges in the northeast (baseline region). 
Each regional coefficient shows the difference from northeast charges. 
Southeast charges are slightly higher, while northwest and southwest are lower, 
though differences are modest.

### Model performance
```{r}
glance_region <- glance(model_region)
glance_region %>% knitr::kable(caption = "Region Model Performance")
```

The model explains less than 1% of variance (R² ≈ 0.006), 
confirming region is not a strong predictor of charges.

## Model 3: Sex + Region

Fit a combined model including both categorical predictors.

```{r}
model_both <- lm(charges ~ sex + region, data = data)
```

### Model results
```{r}
tidy_both <- tidy(model_both, conf.int = TRUE)
tidy_both %>% knitr::kable(caption = "Combined Model Coefficients")
```

In this model, each coefficient represents the effect of that category 
while controlling for the other variable. The sex effect remains similar to the univariate model, 
and regional effects are largely unchanged, suggesting minimal confounding between these variables.

### Model performance
```{r}
glance_both <- glance(model_both)
glance_both %>% knitr::kable(caption = "Combined Model Performance")
```

The combined model explains approximately 1% of variance, only marginally better than either variable alone.

## Model Comparison

### ANOVA comparison
```{r}
anova_sex_both <- anova(model_sex, model_both)
anova_sex_both %>% knitr::kable(caption = "ANOVA: Sex vs Combined")

anova_region_both <- anova(model_region, model_both)
anova_region_both %>% knitr::kable(caption = "ANOVA: Region vs Combined")
```

ANOVA results confirm that adding region to the sex model (or vice versa) 
provides statistically significant improvement, though the practical effect size remains small.

### Information criteria
```{r}
aic_comparison <- AIC(model_sex, model_region, model_both)
aic_comparison %>% knitr::kable(caption = "AIC Comparison")

bic_comparison <- BIC(model_sex, model_region, model_both)
bic_comparison %>% knitr::kable(caption = "BIC Comparison")
```

AIC slightly favors the combined model, while BIC favors the simpler sex-only model. 
This discrepancy reflects BIC's stronger penalty for model complexity. 
Given the minimal improvement in R² when adding region (from 0.5% to 1.1%), 
BIC's preference for the simpler model is more defensible, 
as the added complexity does not justify the marginal gain in explanatory power.

## Changing the Baseline

### Sex with male as baseline
```{r}
data_male_baseline <- data %>%
  mutate(sex = relevel(sex, ref = "male"))

model_sex_male <- lm(charges ~ sex, data = data_male_baseline)
tidy_sex_male <- tidy(model_sex_male, conf.int = TRUE)
tidy_sex_male %>% knitr::kable(caption = "Sex Model (Male Baseline)")
```

When male is the baseline, the intercept now represents average charges for males, 
and the sexfemale coefficient shows females pay $1,387 less. 
The model fit (R²) remains unchanged; only the interpretation shifts.

### Region with southwest as baseline
```{r}
data_sw_baseline <- data %>%
  mutate(region = relevel(region, ref = "southwest"))

model_region_sw <- lm(charges ~ region, data = data_sw_baseline)
tidy_region_sw <- tidy(model_region_sw, conf.int = TRUE)
tidy_region_sw %>% knitr::kable(caption = "Region Model (Southwest Baseline)")
```

With southwest as baseline, all regional coefficients are now relative to southwest charges. 
The model explains the same variance; changing the baseline only alters coefficient interpretation, 
not model quality.

## Manual Dummy Variable Creation

To understand how R handles categorical variables internally, 
we manually create dummy variables and verify the results match automatic factor encoding.

```{r}
data_manual <- data %>%
  mutate(
    # Sex dummy (female = baseline, male = 1)
    male = ifelse(sex == "male", 1, 0),
    
    # Region dummies (northeast = baseline)
    northwest = ifelse(region == "northwest", 1, 0),
    southeast = ifelse(region == "southeast", 1, 0),
    southwest = ifelse(region == "southwest", 1, 0)
  )

model_manual <- lm(charges ~ male + northwest + southeast + southwest, 
                   data = data_manual)
tidy_manual <- tidy(model_manual, conf.int = TRUE)
tidy_manual %>% knitr::kable(caption = "Manual Dummy Variables Model")
```

The manually created dummy variables produce identical results to automatic factor encoding, 
confirming our understanding of R's internal handling of categorical predictors. 
This demonstrates that for k categories, R creates k-1 dummy variables, 
with one category serving as the reference against which others are compared.

## Visualization: Coefficient Plots

### Sex model coefficients
```{r}
tidy_sex %>%
  filter(term != "(Intercept)") %>%
  ggplot(aes(x = term, y = estimate)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  coord_flip() +
  labs(
    x = "Predictor",
    y = "Coefficient Estimate",
    title = "Sex Model: Coefficient with 95% CI"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

The coefficient plot shows males have higher charges than females, 
with the confidence interval clearly excluding zero, indicating statistical significance.

### Region model coefficients
```{r}
tidy_region %>%
  filter(term != "(Intercept)") %>%
  ggplot(aes(x = term, y = estimate)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  coord_flip() +
  labs(
    x = "Predictor",
    y = "Coefficient Estimate",
    title = "Region Model: Coefficients with 95% CI"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

Regional coefficients are relatively small in magnitude and some confidence intervals include zero, 
reflecting weak regional effects on charges.

### Combined model coefficients
```{r}
tidy_both %>%
  filter(term != "(Intercept)") %>%
  ggplot(aes(x = term, y = estimate)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  coord_flip() +
  labs(
    x = "Predictor",
    y = "Coefficient Estimate",
    title = "Combined Model: Coefficients with 95% CI"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

The combined model shows all coefficients remain similar to their univariate counterparts, 
indicating minimal interaction or confounding between sex and region.

## Conclusion

This analysis demonstrates key concepts in handling categorical predictors through regression modeling:

**Dummy Variable Encoding:** R automatically converts categorical variables into dummy variables using treatment coding. 
For a variable with k categories, k-1 dummy variables are created, with one category serving as the reference baseline.

**Reference Category:** By default, R selects the alphabetically first category as the baseline. 
All other categories are compared against this reference through their dummy variable coefficients. 
The baseline can be changed using `relevel()`, which alters interpretation but not model fit.

**Coefficient Interpretation:** The intercept represents the mean outcome for the reference group. 
Each dummy variable coefficient indicates the difference between that category and the reference. 
A positive coefficient means that category has higher charges; negative means lower.

**Model Performance:** Neither sex nor region emerged as strong predictors of insurance charges individually. 
Sex explained approximately 0.5% of variance, region explained 0.6%, and together they explained only 1%. 
This suggests other factors (such as smoking status, age, or BMI) are far more important in determining insurance costs.

**Statistical vs Practical Significance:** While some effects achieved statistical significance due to large sample size, 
the practical impact remains minimal. The difference between males and females ($1,387) is modest compared to 
the range of insurance charges, and regional differences are even smaller.

**Key Takeaway:** Categorical variables in regression require careful interpretation. 
Understanding dummy variable encoding and reference categories is essential for correctly interpreting model coefficients. 
While these models provide a foundation for understanding categorical predictors, 
the weak explanatory power suggests more comprehensive models incorporating additional variables would be necessary 
for meaningful prediction of insurance charges.
